Can we distinguish modes of selective interactions using linkage disequilibrium? ((or)) Mutation correlations in conserved domains reveal a heterogeneous landscape of selective interactions within genes ((or...?))

### 2021-03-01 Manuscript.Rmd
minor edits throughout

Previous theoretical and simulation studies have shown that interference and
interactions between selected mutations reduces the efficacy of selection at
linked loci [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj; @McVean2000-ox],
impacting substitution rates, the deleterious mutation load, and dynamics of
segregating mutations. Interference between moderately deleterious mutations
and synergistic epistasis are both expected to cause negative LD between
selected mutations, which can be readily tested in population genetic data
[@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
numerical approach to generate expectations for signed LD under a wide range of
selective scenarios, and then compared patterns of LD in human populations
between classes of coding mutations within genes using unbiased estimators for
LD from unphased genotypes.

By taking broad genome- or even gene-wide surveys of LD within functional
classes of mutations, patterns of interactive effects that occur within
functional sub-elements of genes can be missed. However, at the gene-wide
level, we already find large differences in the interactions between pairs of
missense and loss-of-function mutations. While LD between pairs of missense
mutations does not appear to be different from pairs of synonymous variants, LD
between loss-of-function variants is substantially more negative.
Hill-Robertson effects under additive selection are expected to be strongest
for slightly to moderately deleterious variants (with $s\sim 1/N_e$), as
strongly deleterious mutations are not expected to interfere with one another
[@McVean2000-ox]. Negative synergistic epistasis between strongly deleterious
mutations does produce large negative deviations of mean LD. However, strongly
deleterious *recessive* mutations can also produce this pattern, making it
difficult to distinguish these effects with certainty. This difficulty is
compounded by the large measurement noise for $\E[D]$, especially for
loss-of-function variants for which only a few hundred pairs are found within
genes in each population.

@@

Positive average LD between missense and synonymous mutations has previously
been reported in humans and other species [@Sohail2017-zq; @Garcia2020-zn;
@Sandler2020-vk], and the similarity of their gene-wide LD might suggest that
interference or interactions between missense mutations are minimal, or at
least no stronger than those between synonymous mutations. However, averaging
over all observed pairs of mutations within a gene completely masks different
interactive effects in functional subunits of genes, effects that even drive LD
in opposite directions. Nonsynonymous mutations found within protein domains,
identified as conserved subunits of a gene with some structural or functional
role, are more strongly selected against on average, but also have increased
signed LD over synonymous mutations at the same distances and also falling
within domains. Missense mutation outside of domains but at the same distances
as those within domains have more negative LD, both compared to synonymous
mutations outside of domains at the same distance and to mutations within
domains. Neither dominance effects, synergistic epistasis, nor Hill-Robertson
interference are expected to result in positive LD, so that some other
interactive effect must be driving this signal within domains.

There are a number of possible interaction scenarios that can result in
positive signed LD between linked loci. One possible explanation is a
prevalence of pairs of compensatory mutations that are tolerated to
co-segregate at high frequencies within conserved domains [@Yeang2007-gj;
@Ivankov2014-tn]. @Callahan2011-ac and @Taverner2020-lk have proposed such a
mechanism to explain observed clusters of mutations in lineages of *Drosophila*
and other species. Another possibility is a model of antagonistic, or
diminishing returns epistasis, in which a single amino acid-changing mutation
within a domain damages the functionality of that subunit, but additional
mutations within that same domain reduces fitness by a factor less than the
first mutation. A third possibility, related to the antagonistic epistasis
case, is that selection acts on the functional domain as a unit instead of at
loci within the domain individually (such as gene-based dominance). In this
scenario, double heterozygotes have different fitnesses depending on whether
the mutations are found on same haplotype or on different haplotypes. Each of
these scenarios lead to positive LD.

@@

## Clustered mutations may drive positive LD among synonymous mutations

In a single randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, we observe
positive LD between synonymous mutations. While selection on some subset of
synonymous variants is possible, it should be much weaker on average than
between missense mutations, and any interference between selected synonymous
variants should lead to negative LD. @Sandler2020-vk proposed that spatial
population structure is responsible for the increase of LD for both synonymous
and missense variants, as every natural population lives in some geography that
will necessarily induce some structure, and they used forward simulations to
demonstrate this effect. However, this increase in LD, at least in simulations,
extends over larger distances separating base-pairs than we
observe in the data. Instead, positive LD between synonymous mutations
decays to zero within just a few hundred base-pairs (Figures \@ref(fig:lddecay)
and \@ref(fig:lddecaydomain)). We might also expect that differences in
migration and admixture histories between human populations would result in
different levels of positive synonymous LD, but the patterns are fairly
consistent across populations.

An alternative explanation for positive LD between neutral variants separated
by short distances is a nonrandom mutational process, where clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of linkage
disequilibrium over short distances, on the order of 10s to 100s of base pairs
[@Harris2014-zg], and clustered mutational events may be common in humans
[@Besenbacher2016-ac]. Thus, positive LD may be the baseline expectation for
tightly linked variants primarily due to clustered mutations, and subsequent
selective processes and interactions will cause LD to deviate from that
expectation, as seen for missense mutations across different sub-elements of
genes. Similarly, it may be more appropriate to compare the negative LD
observed between linked loss-of-function variants to that positive baseline
expectation instead of zero, implying that they would be more recessive or have
stronger synergistic epistasis than if the neutral expectation were zero.

@@

In exploring or inferring selective models across multiple mutations, the
number of plausible selection scenarios becomes quite large as we relax
assumptions about additive and multiplicative fitness functions. The inclusion
of dominance effects, epistasis, or other interactions leads to a rapid
increase in the number of parameters to consider. This makes performing forward
simulations that span the range of all such selective interaction scenarios
burdensome.

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here, allows us
to explore this highly parameterized space of models far more efficiently, and
it opens the possibility for performing likelihood-based inference using signed
LD or other two-locus summaries of the data. For example, inferring the joint
distribution of dominance and selection is underpowered using the SFS. Because
signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
inferring the distribution of fitness effects with dominance may become
feasible using the joint distribution of allele frequencies and LD.

>>>>

Previous theoretical and simulation studies have shown that interference and
interactions between selected mutations reduces the efficacy of selection at
linked loci [@Hill1968-vu; @Birky1988-jm; @Barton1995-mj; @McVean2000-ox],
impacting substitution rates, the deleterious mutation load, and dynamics of
segregating mutations. Interference between moderately deleterious mutations
and synergistic epistasis are both expected to cause negative LD between
selected mutations, which can be readily tested using population genetic data
[@Sohail2017-zq; @Sandler2020-vk; @Garcia2020-zn]. Here, I first used a closed
numerical approach to generate expectations for signed LD under a wide range of
selective scenarios, and then compared patterns of LD in human populations
between classes of coding mutations within genes using unbiased estimators for
LD from unphased genotypes.

By taking broad genome- or gene-wide surveys of LD across functional classes of
mutations, heterogeneous patterns of interactive effects that occur within
functional sub-elements of genes can be missed. However, at the gene-wide
level, we already find large differences in interactions between pairs of
missense and loss-of-function mutations.

From gene-wide averages, LD between pairs of missense mutations does not appear
to be different from pairs of synonymous variants, but LD between
loss-of-function variants is substantially more negative. Hill-Robertson
effects under additive selection are expected to be strongest for slightly to
moderately deleterious variants (with $s\sim 1/N_e$), as strongly deleterious
mutations are not expected to interfere with one another [@McVean2000-ox].
However, loss-of-function mutations do not typically fall within this regime,
and inference of the distribution of fitness effects (DFE) for new
loss-of-function variants shows that a large majority are strongly deleterious
(Supporting Information). Instead, negative synergistic epistasis between
strongly deleterious mutations does produce large negative deviations of mean
LD. Strongly deleterious *recessive* mutations can also produce this pattern,
making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for $\E[D]$, especially
for loss-of-function variants for which only a few hundred pairs are found
within genes in each population. If loss-of-function variants are in fact
recessive, even partially, this can have a large impact on the inferences of
the DFE (Supporting Information).

@@

Positive average LD between missense and synonymous mutations has previously
been reported in humans and other species [@Sohail2017-zq; @Garcia2020-zn;
@Sandler2020-vk], and the similarity of their gene-wide LD might suggest that
interference or interactions between missense mutations are minimal, or at
least no stronger than those between synonymous mutations. However, averaging
over all observed pairs of mutations within a gene masks element-specific
interactive effects that drive LD in opposite directions. Nonsynonymous
mutations found within protein domains, identified as conserved subunits of a
gene with some structural or functional role, are more strongly selected
against on average, but also have increased signed LD over synonymous mutations
at the same distances within domains. Missense mutation outside of domains but
at the same distances as those within domains have more negative LD, both
compared to distance-matched synonymous mutations outside of domains and to
mutations within domains. Neither dominance effects, synergistic epistasis, nor
Hill-Robertson interference are expected to result in positive LD, so that some
other interactive effect must be driving this positive LD signal.

There are a number of possible interaction scenarios that can result in
positive signed LD between linked loci. One explanation is a
prevalence of pairs of compensatory mutations that are tolerated to
co-segregate at high frequencies within conserved domains [@Yeang2007-gj;
@Ivankov2014-tn]. @Callahan2011-ac and @Taverner2020-lk have proposed such a
mechanism to explain observed clusters of mutations in lineages of *Drosophila*
and other species. Another possibility is a model of antagonistic, or
diminishing returns epistasis, in which a single amino acid-changing mutation
within a domain damages the functionality of that subunit, but additional
mutations within that same domain reduces fitness by a factor less than the
first mutation. A third possibility, related to the antagonistic epistasis
case, is that selection acts on the functional domain as a unit instead of at
loci within the domain individually (such as gene-based dominance). In this
scenario, double heterozygotes have different fitnesses depending on whether
the mutations are found on same haplotype or on different haplotypes.

@@

## Clustered mutations can account for positive LD among synonymous mutations

In a single randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, we observe
positive LD between synonymous mutations. While selection on some subset of
synonymous variants is possible, it should be much weaker on average than
between missense mutations, and any interference between selected synonymous
variants should lead to negative LD. @Sandler2020-vk proposed that spatial
population structure is responsible for the increase of LD for both synonymous
and missense variants, arguing that every natural population lives in some
geography that will necessarily induce some structure, and they used forward
simulations to demonstrate this effect. However, this increase in LD, at least
in simulations of spatial structure and migration, typically extends over larger
distances separating base-pairs than we observe in the data. Instead, positive
LD between synonymous mutations decays to zero within just a few hundred
base-pairs (Figures \@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).

Nonrandom mutational processes provide an alternative explanation for positive
LD between neutral variants separated by short distances, in which clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of LD over short
distances, on the order of 10s to 100s of base pairs [@Harris2014-zg], and
clustered mutational events may be common in humans [@Besenbacher2016-ac].
Thus, positive LD may be the baseline expectation for tightly linked variants
primarily due to clustered mutations, and only subsequent selective processes and
interactions will cause LD to deviate from that expectation, as seen for
missense mutations across different sub-elements of genes. Similarly, it may be
more appropriate to compare the negative LD observed between linked
loss-of-function variants to that positive baseline expectation instead of
zero, implying that they would be more recessive or have stronger synergistic
epistasis than if the neutral expectation were zero.

@@

In exploring or inferring multi-locus selective models, the number of plausible
selection scenarios becomes quite large as we relax the strict assumptions of
additive and multiplicative fitness effects. The inclusion of dominance
effects, epistasis, or other interactions leads to a rapid increase in the
number of parameters to consider. This makes performing forward simulations
that span the range of all such selective interaction scenarios burdensome.

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here, allow us
to explore this highly parameterized space of models far more efficiently, and
it opens the possibility for performing likelihood-based inference using signed
LD or other two-locus summaries of the data. For example, inferring the joint
distribution of dominance and selection is underpowered using the SFS. Because
signed LD is sensitive to the levels of dominance (Figure \@ref(fig:fig3)),
inferring the DFE with dominance may become feasible using the joint
distribution of allele frequencies and LD.
